Financialisation of Commodity Markets

### 2024-10-23 docs/paper.qmd
Editorial polishing

---
title: Financialisation of Commodity Markets
subtitle: Co-movement Behind-the-Scenes
author:
  - name: Devraj Basu
    email: devraj.basu@strath.ac.uk
    affiliations: 
        - id: strathclyde
          name: University of Strathclyde Business School
          department: Accounting & Finance
          address: 199 Cathedral Street
          city: Glasgow
          state: Scotland
          postal-code: G4 0QU
    attributes:
        corresponding: true
  - name: Olivier Bauthéac
    email: olivier.bautheac@strath.ac.uk
    affiliations:
        - ref: strathclyde
abstract: |
  In the early 2000s institutional investors entered the commodity futures markets en masse with passive, long only, index type positions in sharp contrast with those typically assumed by traditional expert participants. A heated public debate soon erupted over the perceived consequences of the phenomenon---commonly referred to as "financialisation"---and, in response to immediate policy concerns, the matter was thrust onto the academic sphere as a burning issue. With the benefit of hindsight it now seems that the academic debate was framed rather narrowly, used contentious research methods, and eventually led to regulatory changes that were therefore perhaps unwarranted. In contrast, we take a broader approach where we consider a large cross-section of liquid commodities as suggested by the nature of financialisation that comprehends commodity futures as an asset class. Using a bespoke asset pricing factors based framework we study the issue through the lens of co-movement and find that the phenomenon had ontological consequences for the commodity complex with its impact extending beyond the mechanical effects induced by indexation. The onset of the financial crisis and the monetary policy regimes that followed, on the other hand, seem to have set off a motion of reversion to legacy pre-financialisation fundamentals.
keywords: 
  - Commodity markets
  - Financialisation
  - Asset pricing
date: last-modified
bibliography: bibliography.bib
code_repository: https://github.com/bautheac/financialisation-co-movement
format:
  elsevier-pdf:
    keep-tex: true
    journal:  
      name: Journal of Futures Markets
      formatting: preprint
      model: 3p
      cite-style: authoryear
  html:
    grid:
      body-width: 1000px
    toc: true
    toc_depth: 3
    toc-title: Contents
    css: style.css
    code-links:
      - text: github repository
        icon: github
        href: https://github.com/bautheac/financialisation-co-movement
---

@@

The last twenty years have seen major upheavals in the commodity markets with the onset of financialisation^[Starting around 2004, with a view of commodity markets as an emerging asset class and in an effort to hedge against inflation and seek diversification benefits [@buyuksahin_speculators_2014; @singleton_investor_2013], institutional investors sent forth an unprecedented flow of capital into commodity futures markets. This phenomenon is commonly referred to as the financialisation of commodities [@domanski_financial_2007].] as well as the period of the 2008 financial crisis and its aftermath. Financialisation appears to have affected multiple economic sectors, including agriculture and energy, and its impact has been hotly debated in the policy, legislative, regulatory, and academic spheres. A rigorous legislative scrutiny accompanied by a vigorous policy and academic debate across a number of disciplines about the perceived effects on commodity prices of the unprecedented inflow of institutional funds brought into commodity futures markets by new index type investors^[See section @sec-background for a detailed discussion.] eventually led to regulatory changes in several individual markets.  
In response to immediate policy concerns, the academic debate was initially framed rather narrowly around adequacy of speculation---the burning issue of the day---at the individual commodity market level and the ensuing academic analysis focused on the more mechanical effects of financialisation. With the benefit of hindsight, this approach seems to lack a fundamental understanding of the phenomenon it endeavoured to study and which consequences it attempted to address.  
Financialisation was a phenomenon global in nature that, through the channel of index investment, affected the whole cross-section of liquid commodities at once, eventually spurring its transition away from a collection of idiosyncratic markets to its emergence as an asset class. Focusing on individual markets in this context thus seems limited. Most of the studies underpinning the regulatory response to financialisation took the form of pairwise correlation, causation and to some extent co-integration analysis, all of which are idiosyncratic in nature for they study single pairs of individual assets. Besides, some of these techniques since proved controversial and seem to produce results the quality of which may fall short of the standards that may reasonably be expected to lay the ground for legislative action.  Although appealing, for it comes with the promise of prompt "significant" results, this approach misses a critical corollary of the substantial advent of a new category of players in a (set of) market(s): the alteration of fundamentals it potentially induces, best observed in this context through the lens of commercial hedging pressure [@keynes_treatise_1930; @hicks_value_1939].  
While essential to the understanding of the phenomenon, the global nature of financialisation and its impact on market fundamentals therefore appear to have been originally overlooked and another approach, perhaps more refined, seems warranted.

@@

These entered the commodity spectrum en masse with across-the-board index type positions propping the transition of the complex from a collection of idiosyncratic markets towards a more congruent ensemble akin to an asset class. In this context of heightened cross-asset market participation the issue of interconnectedness stands out as paramount and we strive to investigate it by first examining the correlation structure of our sample set of assets at the global, country, sector and subsector levels. Our results indicate higher interdependence within the full cross-section of commodities over the financialisation period, consistent with both the empirical and theoretical literature. The increase appears to be driven by the US metals sector, a new observation to the best of our knowledge, and, unlike individual commodity returns, does not seem to be regime-driven, as a similar pattern of results is observed across phases of **backwardation** and **contango**.

@@

The returns pull-back continues into the post-crisis period, interestingly amidst a lower volatility environment, suggesting that investment outflows from the commodity futures complex have had a lower impact on volatility than the corresponding inflows witnessed during financialisation. The pattern of reversion toward the pre-financialisation Keynesian paradigm continues stronger with mean returns higher during phases of **backwardation** relative to **contango** and the difference more pronounced than over the crisis period. In contrast to the crisis period when they seemed to be converging, the pricing performance of CHP and **CHP** now part way, seemingly reverting to the Keynesian paradigm observed pre-financialisation, further suggesting a reversion to legacy commodity specific dynamics. The pull back extends to the correlation structure with an across-the-board decrease in average pairwise correlations down to levels below those observed during financialisation. The average market factor beta for the entire cross-section shows a similar pattern, dropping down below the level observed during financialization with the decline seemingly driven by the metals sector, particularly the GB traded segment. The pattern extends to the pricing performance of all factors that drops below levels observed during financialisation while the gap between the legs of the long-short factors remains narrower than observed during that period.  
Moreover, both the average correlations and the pricing performance of the market factor are considerably stronger during **contango** phases, unlike previous periods, which showed no substantial difference between regimes with especially pronounced for metal assets globally, both US and GB traded. These results document the emerging role/dominance of the **contango** regime post-crisis and could provide a valuable basis for further research.

@@

A world-wide debate ensued about the role of index funds in commodity markets. The first responses to the 2007/2008 crisis of escalating food and energy prices took the form of policy reports, many of which reasoned that the growth of commodity index funds came along with an influx of largely speculative capital that was responsible for driving commodity prices beyond their historic highs [@deschutter_food_2010; @gilbert_speculative_2010; @herman_not_2011; @schumann_hunger_2011; @unctad_global_2009].  
Early US Senate investigations on the matter drew on pricing and trading data supplied by the CFTC as well as interviews with numerous experts including hedge fund manager Michael Masters who linked the growing presence of index investors in US agricultural and energy markets to the observed surges in both futures and cash prices [@masters_testimony_2008; @masters_accidental_2008]. This contention has since come to be known as the Masters’ hypothesis and was largely endorsed by the final US Senate report on the issue [@senate_excessive_2009].  
At this stage, the question thrust on the academic sphere was whether "excessive speculation" (understood as the market activity peculiar to index type investors) was linked to escalating energy and agricultural commodity prices. A number of ensuing academic studies, most of which used commodity specific correlation and causality based analysis have since disproved this contention: @irwin_devil_2009, @sanders_adequacy_2010, @sanders_impact_2011, @sanders_new_2011, @irwin_commodity_2013, @brunetti_commodity_2014, @hamilton_effects_2015 and @bruno_financialisation_2017 investigated the issue in the context of the agricultural markets; meanwhile, @buyuksahin_speculators_2011, @tokic_speculation_2012, @fattouh_role_2013, @kilian_role_2014, @knittel_simple_2016 and @manera_modelling_2016 examined the energy markets while @bohl_does_2013, @kim_does_2015 and @boyd_prevalence_2016 studied both energy and agricultural markets and @irwin_index_2011, @irwin_financialisation_2012 as well as @stoll_commodity_2011 examined the commodity markets in general. Other studies have underlined various alterations in commodity pricing dynamics over the period. @gilbert_role_2014 reject the view that financialisation has not had any effects in the grains markets and demonstrate that trades originated by financial market participants, and specifically index investors, can move prices but tend to be typically volatility-reducing. @juvenal_speculation_2015 contend that the oil price increase between 2004 and 2008 was mainly driven by the strength of global demand but that the financialisation process of commodity markets also played a role. Likewise, @henderson_new_2015 show that non-information-based financial investments have important impacts on commodity prices. @cheng_financialisation_2014, on the other hand, show that in the case of index commodities, financialisation has transformed the risk sharing, and information discovery functions of commodity futures markets.

@@

For each period, the return series for the individual US-traded assets are regressed on both the relative change on their respective CHP series and that on the cross-sectional CHP independently. The resulting $R^{2}$ values are then averaged by sector and subsector. Additionally, pairwise correlations are computed for each combination of assets across the periods of interest and for each individual year within the full sample period. These correlation coefficients are subsequently averaged across the entire cross-section of assets, as well as by country, sector, and subsector. For each period, the coefficients are also calculated and averaged across phases of low and high cross-sectional CHP.

@@

@tbl-stats-combined displays descriptive statistics for the commodity series considered in the study; the results show a clear pattern of increase in average return during financialisation with an equally weighted portfolio of the US traded assets showing a mean returns of 15.9% significant at the 5% level vs. 6.9% with a 10% significance in the pre-financialisation period. The increase is even more pronounced for an equally weighted portfolio of GB traded assets (metals), 2.47% (not significant) to 21.8% (significant at the 10% level), which sets out the global nature of the phenomenon. Most commodities show a higher mean return while five---Lumber (XCME), Natural gas (XNYM), Palladium (XNYM), Platinum (XNYM) and Nickel-primary (XLME)---show a lower figure with five significant figures---Crude oil-WTI (XNYM), Heating oil (XNYM), Copper (XCEC), Copper (XLME) and Tin-refined (XLME)---against none pre-financialisation. Sugar-#11 wins the race with a mean return increasing from -4.6% to 24.9% followed by copper with sizeable increases for both markets: 1.5% to 28.2% (XCEC) and 0.5% to 28.6% (XLME).  

@@

The onset of financialisation thus seems to have engendered a change in the nature of hedger’s behaviour, at the aggregate level in particular, taking it away from the traditional Keynesian view of risk shifting. This phenomenon has been noted in earlier theoretical and empirical work [@danthine_information_1978; @stout_why_1998] in different contexts from which two competing models of behaviour have emerged.  
The information arbitrage model [@grossman_impossibility_1980; @danthine_information_1978; @kyle_informed_1989] implies that hedgers may often be seeking out counterparties to trade with rather than being purely passive. This issue is also raised in both @cheng_financialisation_2014 and @stulz_rethink_1996 who point out that hedgers may be taking a view on prices just as speculators do. In fact, by hedging away some of their risk, hedgers are able to speculate more heavily based on their disagreement against speculators regarding futures price movement [@simsek_speculation_2013], which fits into a heterogeneous expectation theory of speculation^[@hirshleifer_speculation_1975; @hirshleifer_reply_1976; @hirshleifer_theory_1977]. As @stout_why_1998 points out, this disagreement-based theory of trading is one of the main reason the public and the law disapproves speculation as this form of trading is regarded as non-productive^[@duffie_challenges_2014 also discusses some of the challenges faced by a policy treatment of speculative trading motivated by differences in beliefs.].  
In the second model hedgers are seen as liquidity providers as in @kang_tale_2020 with this explanation also consistent with the nature of financialisation which led to the arrival of long term, long only investors who require substantial liquidity to roll-over their positions. @tbl-time-reg-US-CHP supports this view. We regress commodity futures front returns on individual commercial hedging pressure (CHP) as well as aggregate CHP (**CHP**), our market wide measure of CHP, in separate models and find that the explanatory power of CHP at the individual commodity level decreases over the financialisation period with the US cross-sectional average $R^{2}$ falling down to 20.8% from 25.8% in the previous period while that of **CHP** increases (7.1% vs 5.4%). Financialisation therefore seems to have had the effect of loosening the extent to which futures risk premiums relate to commodity specific commercial hedging motives while strengthening that to which they relate to broader dynamics such as cross-sectional liquidity provision. The pattern is strongest for the metals sector where the average $R^{2}$ drops down to 19.1% from 28.1% between the two periods at the individual level while it rises from 8.4% to 15.7% at the aggregate level. The symmetry breaks in the crisis period, with the US cross-sectional average $R^{2}$ at the individual level bouncing back up (22.5% vs 20.8%) while that at the aggregate level rises further (14.7% vs 7.1%), before eventually resuming post-crisis with a further rise at the individual level (27.2% vs 22.5%) and a drop at the aggregate level (7.5% vs 14.7%) suggesting a return to legacy commodity specific dynamics.

@@

Financialisation was a global phenomenon, through the channel of index investment it simultaneously influenced the entire spectrum of liquid commodities; this sweeping impact spurred the transformation of commodity markets from a collection of idiosyncratic markets into a unified asset class with interconnectedness emerging as a crucial aspect that warrants thorough examination.   
We strive to address this issue by first examining the average of weekly returns pairwise correlations between commodities across the entire cross-section of the commodity assets considered in the study as well as by country, sector and sub-sector. @tbl-correlations-periods shows the results for the four periods of interest with the results reported for the entire period as well as for periods of **backwardation** and **contango**. For the entire sample the average pairwise correlation jumps from 0.07 in the first period to 0.16 over financialisation. This finding indicate higher interdependence within the full cross-section of commodities starting with the financialisation period, consistent with the empirical findings in @fryMcKibbin_evolution_2023 and @mayer_financialization_2017 as well as with the theoretical model of @basak_model_2016. The sector analysis sheds light on the driving force behind these changes. The US metals see the sharpest increase in pairwise correlations going from 0.22 over the past period to 0.56 over financialisation while the three other sectors see a much smaller percentage increase during financialisation relative to the past period. These results thus suggest that the US metals sector was the driving force behind increased average pairwise correlations during financialisation, a new observation to the best of our knowledge. The pattern of correlations is very similar across phases of **backwardation** and **contango** for all the sectors and sub-sectors over the past and financialisation periods which indicates that increased commodity interdependence over financialisation was not a phase driven phenomenon in the same way that individual commodity returns are, as demonstrated in @tbl-stats-combined.   
We further examine these correlations by year in @tbl-correlations-years to uncover any distinct breakpoints. For the full sample of assets, the average pairwise correlations increase from 0.08 to 0.12 from 2003 to 2004 seemingly driven by US metals average pairwise correlations, more than doubling from 0.22 to 0.49, suggesting 2004 as the onset of financialization as noted in several studies [@baker_financialization_2021; @tang_index_2012] and our contribution is to observe that the US metals sub-sector seems to have played a major role in this.

@@

We investigate further the issue of co-movement from an asset pricing perspective using a bespoke asset pricing framework starting with a market factor mimicking portfolio constituted of the twenty-four US traded commodities considered in the study with equal weighting. This portfolio enables a refined examination of cross correlations across sectors as the numerator of the beta of the regression of the returns for an individual commodity against those for the market factor is driven by covariances while the denominator is driven by the individual commodity variance. Thus, an increase in a commodity's correlation with other commodities should lead to an increase in its beta relative to the market factor portfolio. We accordingly regress the weekly individual commodity front futures returns against those for our market factor portfolio and average the resulting betas across the entire set of commodity assets considered as well as by country, sector and sub-sector for each of the four periods of interest with the results shown in @tbl-regressions-index. The average beta for the US metals sector jumps sharply to 1.12 over financialisation from 0.63 over the previous period; the increase is even more dramatic for the base metals sub-sector (GB traded assets) in a relative sense. The group had a much lower beta of 0.27 over the past period, suggesting a low level of correlation with US commodities; this figure increases to 0.85 over the financialisation period, even beyond that for the US agricultural sector of 0.80. In contrast, the corresponding figure for the energy sector show a substantial decrease, down to 1.61 during financialisation from 2.33 over the past period.   

>>>>

---
title: Financialisation of Commodity Markets
subtitle: Co-movement Behind-the-Scenes
author:
  - name: Devraj Basu
    email: devraj.basu@strath.ac.uk
    affiliations: 
        - id: strathclyde
          name: University of Strathclyde Business School
          department: Accounting & Finance
          address: 199 Cathedral Street
          city: Glasgow
          state: Scotland
          postal-code: G4 0QU
    attributes:
        corresponding: true
  - name: Olivier Bauthéac
    email: olivier.bautheac@strath.ac.uk
    affiliations:
        - ref: strathclyde
abstract: |
  In the early 2000s institutional investors entered the commodity futures markets en masse with passive, long only, index type positions in sharp contrast with those typically assumed by traditional expert participants. A heated public debate soon erupted over the perceived consequences of the phenomenon---commonly referred to as "financialisation"---and, in response to immediate policy concerns, the matter was thrust onto the academic sphere as a burning issue. With the benefit of hindsight it now seems that the academic debate was framed rather narrowly, used contentious research methods, and eventually led to regulatory changes that were therefore perhaps unwarranted. In contrast, we take a broader approach where we consider a large cross-section of liquid commodities as suggested by the nature of financialisation that comprehends commodity futures as an asset class. We examine the cross-sectional interconnectedness of these assets and uses a bespoke asset pricing factors based framework to study the issue through the lens of co-movement. We find that the phenomenon had ontological consequences for the commodity complex with its impact extending beyond the mechanical effects induced by indexation. The onset of the financial crisis and the monetary policy regimes that followed, on the other hand, seem to have set off a motion of reversion to legacy pre-financialisation fundamentals.
keywords: 
  - Commodity markets
  - Financialisation
  - Asset pricing
date: last-modified
bibliography: bibliography.bib
code_repository: https://github.com/bautheac/financialisation-co-movement
format:
  elsevier-pdf:
    keep-tex: true
    journal:  
      name: Journal of Futures Markets
      formatting: preprint
      model: 3p
      cite-style: authoryear
  html:
    grid:
      body-width: 1000px
    toc: true
    toc_depth: 3
    toc-title: Contents
    css: style.css
    code-links:
      - text: github repository
        icon: github
        href: https://github.com/bautheac/financialisation-co-movement
---

@@

The last twenty years have seen major upheavals in the commodity markets with the onset of financialisation^[Starting around 2004, with a view of commodity markets as an emerging asset class and in an effort to hedge against inflation and seek diversification benefits [@buyuksahin_speculators_2014; @singleton_investor_2013], institutional investors sent forth an unprecedented flow of capital into commodity futures markets. This phenomenon is commonly referred to as the financialisation of commodities [@domanski_financial_2007].] as well as the period of the 2008 financial crisis and its aftermath. Financialisation appears to have affected multiple economic sectors, including agriculture and energy, and its impact has been hotly debated in the policy, legislative, regulatory, and academic spheres. A rigorous legislative scrutiny accompanied by a vigorous policy and academic debate across a number of disciplines about the perceived effects on commodity prices of the unprecedented inflow of institutional funds brought into commodity futures markets by new index type investors^[See section @sec-background for a detailed discussion.] eventually led to regulatory changes in several individual markets.  
In response to immediate policy concerns, the academic debate was initially framed rather narrowly around adequacy of speculation---the burning issue of the day---at the individual commodity level and the ensuing academic analysis focused on the more mechanical effects of financialisation. With the benefit of hindsight, this approach seems to lack a fundamental understanding of the phenomenon it endeavoured to study and which consequences it attempted to address.  
Financialisation was a phenomenon global in nature that, through the channel of index investment, affected the whole cross-section of liquid commodities at once, eventually spurring its transition away from a collection of idiosyncratic markets to its emergence as an asset class. Focusing on individual markets in this context thus seems limited. Most of the studies underpinning the regulatory response to financialisation took the form of pairwise correlation, causation and to some extent co-integration analysis, all of which are idiosyncratic in nature for they study single pairs of individual assets. Besides, some of these techniques since proved controversial and seem to produce results the quality of which may fall short of the standards that may reasonably be expected to lay the ground for legislative action.  Although appealing, for it comes with the promise of prompt "significant" results, this approach misses a critical corollary of the substantial advent of a new category of players in a (set of) market(s): the alteration of fundamentals it potentially induces, best observed in this context through the lens of commercial hedging pressure [@keynes_treatise_1930; @hicks_value_1939].  
While essential to the understanding of the phenomenon, the global nature of financialisation and its impact on market fundamentals therefore appear to have been originally overlooked and another approach, perhaps more refined, seems warranted.

@@

These entered the commodity spectrum en masse with across-the-board index type positions propping the transition of the complex from a collection of idiosyncratic markets towards a more homogeneous ensemble akin to an asset class. In this context of heightened cross-asset market participation the issue of interconnectedness stands out as paramount and we strive to investigate it by first examining the correlation structure of our sample set of assets at the global, country, sector and subsector levels. Our results indicate higher interdependence within the full cross-section of commodities over the financialisation period, consistent with both the empirical and theoretical literature. The increase appears to be driven by the US metals sector, a new observation to the best of our knowledge, and, unlike individual commodity returns, does not seem to be regime-driven, as a similar pattern of results is observed across phases of **backwardation** and **contango**.

@@

The returns pull-back continues into the post-crisis period, interestingly amidst a lower volatility environment, suggesting that investment outflows from the commodity futures complex have had a lower impact on volatility than the corresponding inflows witnessed during financialisation. The pattern of reversion toward the pre-financialisation Keynesian paradigm continues stronger with mean returns higher during phases of **backwardation** relative to **contango** and the difference more pronounced than over the crisis period. In contrast to the crisis period when they seemed to be converging, the pricing performance of CHP and **CHP** now part way, seemingly reverting to the Keynesian paradigm observed pre-financialisation, further suggesting a reversion to legacy commodity specific dynamics. The pull back extends to the correlation structure with an across-the-board decrease in average pairwise correlations down to levels below those observed during financialisation. The average market factor beta for the entire cross-section shows a similar pattern, dropping down below the level observed during financialization with the decline seemingly driven by the metals sector, particularly the GB traded segment. The pattern extends to the pricing performance of all factors that drops below levels observed during financialisation.  
Moreover, both the average correlations and the pricing performance of the market factor are considerably stronger during **contango** phases, unlike previous periods, which showed no substantial difference between regimes with the pattern particularly pronounced for metal assets globally, both US and GB traded. These results document the emerging role/dominance of the **contango** regime post-crisis and could provide a valuable basis for further research.

@@

A world-wide debate ensued about the role of index funds in commodity markets. The first responses to the 2007/2008 crisis of escalating food and energy prices took the form of policy reports, many of which reasoned that the growth of commodity index funds came along with an influx of largely speculative capital that was responsible for driving commodity prices beyond their historic highs [@deschutter_food_2010; @gilbert_speculative_2010; @herman_not_2011; @schumann_hunger_2011; @unctad_global_2009].  
Early US Senate investigations on the matter drew on pricing and trading data supplied by the CFTC as well as interviews with numerous experts including hedge fund manager Michael Masters who linked the growing presence of index investors in US agricultural and energy markets to the observed surges in both futures and cash prices [@masters_testimony_2008; @masters_accidental_2008]. This contention has since come to be known as the "Masters’ hypothesis" and was largely endorsed by the final US Senate report on the issue [@senate_excessive_2009].  
At this stage, the question thrust on the academic sphere was whether "excessive speculation" (understood as the market activity peculiar to index type investors) was linked to escalating energy and agricultural commodity prices. A number of ensuing academic studies, most of which used commodity specific correlation and causality based analysis have since disproved this contention: @irwin_devil_2009, @sanders_adequacy_2010, @sanders_impact_2011, @sanders_new_2011, @irwin_commodity_2013, @brunetti_commodity_2014, @hamilton_effects_2015 and @bruno_financialisation_2017 investigated the issue in the context of the agricultural markets; meanwhile, @buyuksahin_speculators_2011, @tokic_speculation_2012, @fattouh_role_2013, @kilian_role_2014, @knittel_simple_2016 and @manera_modelling_2016 examined the energy markets while @bohl_does_2013, @kim_does_2015 and @boyd_prevalence_2016 studied both energy and agricultural markets and @irwin_index_2011, @irwin_financialisation_2012 as well as @stoll_commodity_2011 examined the commodity markets in general. Other studies have underlined various alterations in commodity pricing dynamics over the period. @gilbert_role_2014 reject the view that financialisation has not had any effects in the grains markets and demonstrate that trades originated by financial market participants, and specifically index investors, can move prices but tend to be typically volatility-reducing. @juvenal_speculation_2015 contend that the oil price increase between 2004 and 2008 was mainly driven by the strength of global demand but that the financialisation process of commodity markets also played a role. Likewise, @henderson_new_2015 show that non-information-based financial investments have important impacts on commodity prices. @cheng_financialisation_2014, on the other hand, show that in the case of index commodities, financialisation has transformed the risk sharing, and information discovery functions of commodity futures markets.

@@

For each period, the return series for the individual US-traded assets are regressed on both the relative change on their respective CHP series and that on the cross-sectional CHP independently. The resulting $R^{2}$ values are then averaged by sector and subsector. Additionally, pairwise correlations are computed for each combination of assets across the periods of interest and for each individual year within the full sample period. These correlation coefficients are then averaged across the entire cross-section of assets, as well as by country, sector, and subsector. For each period, the coefficients are also calculated and averaged across phases of low and high cross-sectional CHP.

@@

@tbl-stats-combined displays descriptive statistics for the commodity series considered in the study; the results show a clear pattern of increase in average return during financialisation with an equally weighted portfolio of the US traded assets showing a mean returns of 15.9% significant at the 5% level vs. 6.9% with a 10% significance in the pre-financialisation period. The increase is even more pronounced for an equally weighted portfolio of GB traded assets (metals), 2.47% (not significant) to 21.8% (significant at the 10% level), which sets out the global nature of the phenomenon. Most commodities show a higher mean return while five---Lumber (XCME), Natural gas (XNYM), Palladium (XNYM), Platinum (XNYM) and Nickel-primary (XLME)---show a lower figure with five og them significant---Crude oil-WTI (XNYM), Heating oil (XNYM), Copper (XCEC), Copper (XLME) and Tin-refined (XLME)---against none pre-financialisation. Sugar-#11 wins the race with a mean return increasing from -4.6% to 24.9% followed by copper with sizeable increases for both markets: 1.5% to 28.2% (XCEC) and 0.5% to 28.6% (XLME).  

@@

The onset of financialisation thus seems to have engendered a change in the nature of hedger’s behaviour, at the aggregate level in particular, taking it away from the traditional Keynesian view of risk shifting. This phenomenon has been noted in earlier theoretical and empirical work [@danthine_information_1978; @stout_why_1998] in different contexts from which two competing models of behaviour have emerged.  
The information arbitrage model [@grossman_impossibility_1980; @danthine_information_1978; @kyle_informed_1989] implies that hedgers may often be seeking out counterparties to trade with rather than being purely passive. This issue is also raised in both @cheng_financialisation_2014 and @stulz_rethink_1996 who point out that hedgers may be taking a view on prices just as speculators do. In fact, by hedging away some of their risk, hedgers are able to speculate more heavily based on their disagreement against speculators regarding futures price movement [@simsek_speculation_2013], which fits into a heterogeneous expectation theory of speculation^[@hirshleifer_speculation_1975; @hirshleifer_reply_1976; @hirshleifer_theory_1977]. As @stout_why_1998 points out, this disagreement-based theory of trading is one of the main reason the public and the law disapproves speculation as this form of trading is regarded as non-productive^[@duffie_challenges_2014 also discusses some of the challenges faced by a policy treatment of speculative trading motivated by differences in beliefs.].  
In the second model hedgers are seen as liquidity providers as in @kang_tale_2020 with this explanation also consistent with the nature of financialisation which led to the arrival of long term, long only investors who require substantial liquidity to roll-over their positions. @tbl-time-reg-US-CHP supports this view. We regress commodity futures front returns on individual commercial hedging pressure (CHP) as well as aggregate CHP (**CHP**), our market wide measure of CHP, in separate models and find that the explanatory power of CHP at the individual commodity level decreases over the financialisation period with the US cross-sectional average $R^{2}$ falling down to 20.8% from 25.8% in the previous period while that of **CHP** increases (7.1% vs 5.4%). Financialisation therefore seems to have had the effect of loosening the extent to which futures risk premiums relate to commodity specific commercial hedging motives while strengthening that to which they relate to broader dynamics such as cross-sectional liquidity provision. The pattern is strongest for the metals sector where the average $R^{2}$ drops down to 19.1% from 28.1% between the two periods at the individual level while it rises from 8.4% to 15.7% at the aggregate level.

@@

Financialisation was a global phenomenon, through the channel of index investment it simultaneously influenced the entire spectrum of liquid commodities, eventually spurring the transformation of commodity markets from a collection of idiosyncratic markets into a more homogeneous collection akin to an asset class. In this context, interconnectedness emerges as a crucial aspect that warrants thorough examination.   
We strive to address this issue by first examining the average of weekly returns pairwise correlations between commodities across the entire cross-section of the commodity assets considered in the study as well as by country, sector and sub-sector. @tbl-correlations-periods shows the results for the four periods of interest with the results reported for the entire period as well as for periods of **backwardation** and **contango**. For the entire sample the average pairwise correlation jumps from 0.07 in the first period to 0.16 over financialisation. This finding indicates higher interdependence within the full cross-section of commodities starting with the financialisation period, consistent with the empirical findings in @fryMcKibbin_evolution_2023 and @mayer_financialization_2017 as well as with the theoretical model of @basak_model_2016. The sector analysis sheds light on the driving force behind these changes. The US metals see the sharpest increase in pairwise correlations going from 0.22 over the past period to 0.56 over financialisation while the three other sectors see a much smaller percentage increase during financialisation relative to the past period. These results thus suggest that the US metals sector was the driving force behind increased average pairwise correlations during financialisation, a new observation to the best of our knowledge. The pattern of correlations is very similar across phases of **backwardation** and **contango** for all the sectors and sub-sectors over the past and financialisation periods which indicates that increased commodity interdependence over financialisation was not a phase driven phenomenon in the same way that individual commodity returns are, as demonstrated in @tbl-stats-combined.   
We break down these average correlations by years in @tbl-correlations-years. For the full sample of assets, the average pairwise correlations remain in the 0.06–0.08 range from 1998 through 2003, before suddenly increasing to 0.12 from 2003 to 2004. This suggests 2004 as a distinct break point for the onset of financialisation, as noted in several studies [@baker_financialization_2021; @tang_index_2012]. We further observe that the US metals appear to have played a major role in this as the increase seems driven by the sector which average pairwise correlation more than doubles (0.49 vs 0.22) over the period.

@@

We investigate further the issue of co-movement from an asset pricing perspective using a bespoke asset pricing framework starting with a market factor mimicking portfolio constituted of the twenty-four US traded commodities with equal weighting. This portfolio enables a refined examination of cross-correlations across sectors as the numerator of the beta of the regression of the returns for an individual commodity against those for the market factor is driven by covariances while the denominator is driven by the individual commodity variance. Thus, an increase in a commodity's correlation with other commodities should lead to an increase in its beta relative to the market factor portfolio. We accordingly regress the weekly individual commodity front futures returns against those for our market factor portfolio and average the resulting betas across the entire set of commodity assets considered as well as by country, sector and sub-sector for each of the four periods of interest with the results shown in @tbl-regressions-index. The average beta for the US metals sector jumps sharply to 1.12 over financialisation from 0.63 over the previous period; the increase is even more dramatic for the base metals sub-sector (GB traded assets) in a relative sense. The group had a much lower beta of 0.27 over the past period, suggesting a low level of correlation with US commodities; this figure increases to 0.85 over the financialisation period, even beyond that for the US agricultural sector of 0.80. In contrast, the corresponding figure for the energy sector show a substantial decrease, down to 1.61 during financialisation from 2.33 over the past period.   
The corresponding pricing performance displays a similar pattern as shown in @tbl-regressions-factors where for each period the corresponding $R^{2}$ values are independently averaged over the entire period as well as over periods of **contango** and **backwardation**. The average $R^{2}$ for the market factor across the entire cross-section of commodity assets, including the GB traded base metals, increases to around 20% during financialisation compared to around 10% over the previous period with the rise driven by the metals sector at large. The sector average for the US traded assets jumps to 31% during financialisation from 7.5% over the past period while for the GB traded assets the corresponding figure increases from less than 2%, or virtually no explanatory power for the (US based) market factor, to almost 12% over the financialisation period. The results for our other risk factor mimicking portfolios show a similar pattern and provide a deeper insights into co-movement dynamics in this context. All three factors---CHP, open interest and term structure---performed poorly in the first period, indicating that the commodities with extreme factor attributes (high or low CHP for example) were uncorrelated with the rest of the commodity cross-section, and the overall performance consistent across periods of **contango** and **backwardation**. The onset of financialisation leads to a substantial improvement in the performance of the CHP factor with its average $R^{2}$, albeit still small at 5.4%, almost tripling from the past to the financialisation period in contrast to the other factors that show weaker dynamics with a slight decrease and a much smaller increase for the open interest and term structure factors respectively. Here again the increase in pricing performance for the CHP factor is driven by the metals sector, particularly the precious sub-sector with the average $R^{2}$ for the group jumping to almost 26% during financialisation from 1.2% over the past period. The corresponding figures of 22% and 1.1% for the entire US metals sector stand in sharp contrast with the agricultural sector, which exhibits no notable change, and the energy sector, where pricing performance drops substantially from 9.2% down to 1.6%. The improvement in the CHP factor's pricing performance remarkably comes predominantly from the long leg (low CHP assets: Keynesian "normal backwardation") of the factor^[Results available from the authors upon request.] with the corresponding mimicking portfolio's performance itself driven by the metals, the precious sub-sector in particular. The average $R^{2}$ for the group jumps to 55% during financialisation from less than 12% over the past period while the short leg sees a far smaller increase, up to 8.9% from 3.3% over the past period. Interestingly, although the term structure factor itself only sees a very modest increase, its short leg (low roll-yield assets: Working's backwardation) shows a sizeable rise in pricing performance, yet again driven the metals sector, with the average $R^{2}$ for the group's US traded assets jumping to 28.2% during financialisation from 6.2% over the past period. The pattern of results for both factors extends to the GB traded assets and suggest a convergence of Keneysian and Working's term structure perspective over the financialisation period.   
